Hierarchical time series forecasting in Emergency Medical Services

### 2022-11-14 manuscript/02-literature.qmd
update section 1 and 2

# Research background {#lit}
Emergency medical services (EMS), also referred to as ambulance, paramedic or pre-hospital emergency services, are a critical component in the delivery of urgent medical care to communities. An effective service delivery requires accurate resource planning that are generally replying on demand forecasts at operational, tactical and strategic levels.

There is a substantial literature on the application of forecasting in the Emergency Medical Services. Various areas have been the main focus. One of the major research topics has been forecasting call volume arrivals. @ibrahim2016modeling provide an extensive review of the forecasting models in this context. Another important focus of the literature in EMS is related to forecasting ambulance demand. Although sometimes the definition of the demand is not cleanly stated, however this is generally the demand where a physical resource has been deployed to respond to an incident, which is also referred to attended incidents. Another demand related variable is the verified incidents. These are all incidents that require an action: either send a physical vehicle, deal with via the Clinical Support Desk, get an external (private) provider to respond to it, or send it through to other channels such as police, firefighters or general practitioners. Another domain that has been extensively studied in the literature and it is also close to ours is forecasting Emergency Department attendance. We refer interested readers to @mingliterature2022, @gul2020exhaustive and @wargon2009systematic for extensive reviews of the literature in these areas. Although crucial to EMS performance, @aringhieri2017emergency state that demand forecasting has received limited research attention in the EMS context. In this section, we provide a brief review of studies on forecasting ambulance demand in EMS.

There are generally two main streams of research in this area: i) the first stream is focusing on the application of time series methods and regression approaches on forecasting aggregate ambulance demand as the most widely used approaches [@vile2012predicting; @sasaki2010using]; ii) another stream considers forecasting EMS demand in more finer temporal and geographical granularitties by employing temporal-spatial prediction methods [@zhou2016predicting;@zhou2016predictinglit]. Papers studying the second stream has not used classical regression and time-series approaches like the first one. Our study belongs to the first category, but we briefly review research in both streams.

@sasaki2010using develop a multivariable regression model to estimate future EMS demands. In addition to the historical demand, the population census for different age groups and counts of the number of companies employing more than five people are included in the regression. The census variables describe groups who may be more likely to need an ambulance. A stepwise ordinary least squares regression analysis with SPSS is used for estimating the parameter and generating forecast. The only performance measure reported in this study is $R^2$. The research design of this study is not rigorous and the study is not reproducible. @vile2012predicting explore using a Singular Spectrum Analysis (SSA) method to generate forecasts of the EMS demand at the national level for 7-day, 14-day, 21-day and 28-day forecast horizons using data provided by the Welsh Ambulance Service Trust (WAST). The performance of this approach is compared with Auto-Regressive Moving Average (ARIMA) and Holt-winter time series methods using Root Mean Squared Error (RMSE). They concluded that point forecasts generated by SSA are more accurate for longer-term, but ARIMA and Holt-winter performance is superior for shorter-term horizons. @vile2016time further develop a decision support system to integrate forecasts generated by SSA. However, the study does not compare and contrast of the performance of forecasting methods based on some utility measures such as cost, resource utilization or response time. The tool contains options that allow generating forecasts at various levels of granularity, however, it ignore the hierarchical and grouped relationships structure, preventing  aligned decision making and coordination.

@9659837 investigate forecasting EMS demand in a high spatio-temporal resolution of 1×1km spatial regions and 1-hr time intervals using total incidents in Oslo, Norway from January 1st, 2015 up to and including February 11th, 2019. They used multi-layer perceptron (MLP) and long short-term memory (LSTM) models to forecast the EMS demand, and compare them to simple aggregation methods and baselines. The point forecast accuracy is evaluated using Mean Absolute Error (MAE) and Mean Squared Error (MSE) and the forecast distribution is measures by
Categorical Cross-Entropy. They shows that while Neural Network models perform better in producing point forecast, a distribution baseline method based on spatial distribution of the incidents across all time steps provide more accurate forecast distribution. @zhou2016predictinglit propose three methods based on Gaussian mixture models, kernel density estimation, and kernel warping to predict 4 weeks into future for a 1-km2 spatial region over an hour. Two years of incidents attended from Toronto, Canada (years 2007 and 2008 with 391,296 events) and Melbourne, Australia (years 2011 and 2012 with  696,975 events) are used to build the model and examine the  performance on test data using mean negative log likelihood. They show that forecasts generated by the proposed methods are significantly more accurate than the current industry practice,  a simple averaging formula. @grekousis2019will investigate the combination of spatial analysis methods with data mining techniques based on  an improved Hungarian algorithm and MLP neural network to identify the most likely locations of future emergency events. The proposed approach is tested using data of 2851 events attended by the EMS in Athens, Greece over 24 weeks. They show that 23.24% of real emergency events lie within 50 m of the predicted ones and nearly 70% of the real emergency events lie no further away than 150 m, which is rather accurate given the granularity of the problem at the city level.

One of the major limitations of the current forecasting research in healthcare and EMS is the lack of a forecasting methodology that accounts for the inherent hierarchical and grouped structure in data to produce coherent forecasts, enabling aligned planning and decision making and possibly forecast accuracy improvement by exploring the demand relationships between different levels. While this methodology has been developed over the past 10 years [@panagiotelis2022probabilistic], it has never been explored in this area. Additionally, we found that most research only deals with point forecast estimation and also reproducibility is still a major concern in this area. Also, given the count nature of Considering these limitations from the literature, we develop a forecasting experiment to fill these gaps.

>>>>

# Research background {#lit}
Emergency medical services (EMS) are a critical component in the delivery of urgent medical care to communities. An effective service delivery requires accurate resource planning that are generally replying on demand forecasts at operational, tactical and strategic levels.

There is a substantial number of studies on the application of time series forecasting in the Emergency Medical Services. Various areas have been the focus of the literature. Forecasting call volume arrivals is one of the major research topics. @ibrahim2016modeling provide an extensive review of the forecasting models in this context. Another important area is related to forecasting ambulance demand. Although the definition of demand might not be always clearly stated, however, this is typically referring to a situation where a physical resource has been deployed to respond to an incident. This might be also called _attended incidents_. Another demand related variable is verified incidents. These are all incidents that require an action: either send a physical vehicle, deal with via the Clinical Support Desk (e.g. calls), get an external (private) provider to respond to it, or send it through to other channels such as police, firefighters or general practitioners. Our study is aligned with this stream of the literature. Another similar area that is largely studied in the literature, is Emergency Department forecasting. We refer interested readers to @mingliterature2022, @gul2020exhaustive and @wargon2009systematic for extensive reviews of the literature on Emergency Department forecasting. Although crucial to EMS performance, @aringhieri2017emergency state that demand forecasting has received limited research attention in the EMS context. In this section, we provide a brief review of studies on forecasting ambulance demand in EMS.

There are generally two main streams of research related to forecasting ambulance demand in EMS: i) the first stream focuses on the application of time series methods and regression approaches on forecasting aggregate ambulance demand [@vile2012predicting; @sasaki2010using]; ii) the second stream considers forecasting EMS demand in a more finer temporal and geographical granularitties by employing temporal-spatial prediction methods [@zhou2016predicting;@zhou2016predictinglit]. The focus of our study is related to the first stream of research.

@sasaki2010using develop a multivariable regression model to estimate future EMS demands. In addition to the historical demand, the population census for different age groups and counts of the number of companies employing more than five people are included in the regression. The census variables describe groups who may be more likely to need an ambulance. A stepwise ordinary least squares regression analysis with SPSS is used for estimating the parameter and generating forecast. The only performance measure reported in this study is $R^2$. The research design of this study is not rigorous and the study is not reproducible. @vile2012predicting explore using a Singular Spectrum Analysis (SSA) method to generate forecasts of the EMS demand at the national level for 7-day, 14-day, 21-day and 28-day forecast horizons using data provided by the Welsh Ambulance Service Trust (WAST). The performance of this approach is compared with Auto-Regressive Moving Average (ARIMA) and Holt-winter time series methods using Root Mean Squared Error (RMSE). They concluded that point forecasts generated by SSA are more accurate for longer-term, but ARIMA and Holt-winter performance is superior for shorter-term horizons. @vile2016time further develop a decision support system to integrate forecasts generated by SSA. However, the study does not compare and contrast the performance of forecasting methods based on utility measures such as cost, resource utilization or response time. The tool contains options that allow generating forecasts at various levels of granularity, however, it ignore the hierarchical and grouped relationships structure, preventing  aligned decision making and coordination.

@9659837 investigate forecasting EMS demand in a high spatio-temporal resolution of 1×1km spatial regions and 1-hr time intervals using total incidents in Oslo, Norway from January 1st, 2015 up to and including February 11th, 2019. They used multi-layer perceptron (MLP) and long short-term memory (LSTM) models to forecast the EMS demand, and compare them to simple aggregation methods and baselines. The point forecast accuracy is evaluated using Mean Absolute Error (MAE) and Mean Squared Error (MSE) and the forecast distribution is measures by Categorical Cross-Entropy. They shows that while Neural Network models perform better in producing point forecast, a distribution baseline method based on spatial distribution of the incidents across all time steps provide more accurate forecast distribution. @zhou2016predictinglit propose three methods based on Gaussian mixture models, kernel density estimation, and kernel warping to predict 4 weeks into future for a 1-km2 spatial region over an hour. Two years of incidents attended from Toronto, Canada (years 2007 and 2008 with 391,296 events) and Melbourne, Australia (years 2011 and 2012 with  696,975 events) are used to build the model and examine the  performance on test data using mean negative log likelihood. They show that forecasts generated by the proposed methods are significantly more accurate than the current industry practice,  a simple averaging formula. @grekousis2019will investigate the combination of spatial analysis methods with data mining techniques based on  an improved Hungarian algorithm and MLP neural network to identify the most likely locations of future emergency events. The proposed approach is tested using data of 2851 events attended by the EMS in Athens, Greece over 24 weeks. They show that $23.24\%$ of real emergency events lie within 50 meter of the predicted ones and nearly $70\%$ of the real emergency events lie no further away than 150 meter, which is rather accurate given the granularity of the problem at the city level.

We observed a number of limitations in the literature of EMS forecasting, that encourage us to undertake this research. These limitations are summarized as following:

1. Current studies ignore the inherent hierarchical and/or grouped structure of the time series data and the relationship between series at different levels of hierarchy. This may result in incoherent forecasts leading to misaligned planning and decision making. While the hierarchical forecasting methodology has been developed and applied in various domains over the past 10 years [@panagiotelis2022probabilistic], it has never been explored in this area.

2. Current research is mainly concerned with generating point forecast at a single level of hierarchy. There is a lack of studies presenting the entire forecast distribution of daily ambulance demand for the entire hierarchy to inform the whole decision-making process and to better represent the uncertainty of future demand, providing a risk management tool for planners.

3. Reproducibility is still a major challenge in EMS forecasting, as it is unlikely to reproduce the results without the help of the authors of those papers.

4. Another limitation is related to the generated forecasts that are not integer counts. Since actual ambulance counts cannot be negative or fractions, ambulance demand forecasts should be the same. While this might not be an issue when producing forecasts at a single level, producing non-negative count forecasts in a hierarchical/grouped structure is challenging and requires further investigation in the future.

This paper concerns the problem of hierarchical forecasting in EMS and generates and evaluates both point and probabilistic forecast across different levels of the hierarchy, hence addressing some important gaps identified in the literature.

### 2022-11-14 manuscript/main.qmd
update section 1 and 2

---
title: "Hierarchical and grouped time series forecasting: an application to Emergency Medical Services"
execute:
  echo: false
  message: false
  warning: false
  cache: true
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Annals of Emergency Medicine
      formatting: preprint
      model: 3p
      cite-style: authoryear
  elsevier-html: 
    toc: true
author:
  - name: Bahman Rostami-Tabar
    email: rostami-tabarb@cardiff.ac.uk
    affiliations: 
        - id: carbs
          name: Cardiff University
          department: Cardiff Business School
          country: United Kingdom
          postal-code: CF10 3EU
    attributes:
        corresponding: true
  - name: Rob J. Hyndman
    email: Rob.Hyndman@monash.edu
    affiliations:
        - id: monash
          name: Monash University
          department: Department of Econometrics and Business Statistics
          country: Australia
          postal-code: VIC 3800
abstract: |
  The primary goal of an emergency service is to minimize the ambulance response time. Accurate forecasts of ambulance demand are crucial inputs into the decisions of planners and policymakers on the resource (e.g. staff and fleet) planning and dynamic deployment. Informing those decisions require detailed forecasting of demand at multiple levels of hierarchy from the national and regional levels to the nature of incidents and their priorities. These forecasts are often generated independently by different teams. However, independent forecasts may not add up to the other levels resulting in mismatched forecasts that lead to conflicting decisions and a lack of coordination in the care service. To address this issue, we explore the inherently hierarchical and grouped structure of Emergency Medcine Services. We apply forecast reconciliation methods to generate both point and probabilistic forecasts that are coherent, for the first time in the healthcare literature.
keywords: [Emergency services, healthcare, hierarchical time series forecasting, forecast reconciliation]
bibliography: mybibfile.bib
---

>>>>

---
title: "Hierarchical time series forecasting in Emergency Medical Services"
author:
  - name: Bahman Rostami-Tabar
    email: rostami-tabarb@cardiff.ac.uk
    affiliations: 
        - id: carbs
          name: Cardiff University
          department: Cardiff Business School
          country: United Kingdom
          postal-code: CF10 3EU
    attributes:
        corresponding: true
  - name: Rob J. Hyndman
    email: Rob.Hyndman@monash.edu
    affiliations:
        - id: monash
          name: Monash University
          department: Department of Econometrics and Business Statistics
          country: Australia
          postal-code: VIC 3800
abstract: |
  The primary goal of an emergency service is to minimize the ambulance response time. Accurate forecasts of ambulance demand are crucial inputs into the decisions of planners and policymakers on the resource (e.g. staff and fleet) planning and dynamic deployment. Informing those decisions require detailed forecasting of demand at multiple levels of hierarchy from the national and regional levels to the nature of incidents and their priorities. These forecasts are often generated independently by different teams. However, independent forecasts may not add up to the other levels resulting in mismatched forecasts that lead to conflicting decisions and a lack of coordination in the care service. To address this issue, we explore the inherently hierarchical and grouped structure of Emergency Medcine Services. We apply forecast reconciliation methods to generate both point and probabilistic forecasts that are coherent, for the first time in the healthcare literature.
keywords: 
  - forecasting
  - healthcare
  - emergency services
  - forecast reconciliation
  - hierarchical time series
  - ambulance demand
  - atteneded incidents

bibliography: bibliography.bib
execute:
  echo: false
  message: false
  warning: false
  cache: true
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Annals of Emergency Medicine
      formatting: preprint
      model: 3p
      cite-style: authoryear
  elsevier-html: 
    toc: true
---
